Multidimensional analysis and detection of informative features in human brain white matter

### 2021-04-20 reports/manuscript/paper.tex
Add tractography details

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{1,4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, WA, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, USA
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, WA, USA
\\
\bigskip

@@

samples along the length of each bundle \replaced{\mbox{{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}}}}{\mbox{\cite{yeatman2012tract,
colby2012, ODonnell2009-uu}}},

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
Unlike many previous analyses of white matter (e.g., using TBSS \cite{smith2006tract} and connectometry \cite{yeh2016connectometry}), which focus on classical inference of the differences between groups or individuals in terms of their white matter properties, the focus of the present work is on the combination of features that facilitate accurate prediction of the individual differences in a \emph{phenotype}.
For example, an accurate prediction of which individuals in a group have a particular disease, or a prediction of their age with small error. The distinct goals of \emph{inference} and \emph{prediction} are not necessarily incongruent, but can sometimes be in tension \cite{Bzdok2020-sd}. This distinction is also similar to the distinction between ``encoding'' and ``decoding'' used in the functional MRI literature \cite{Naselaris2011-ss}. In the present work, we predict the phenotypical variance in a group of subjects (or classify group membership), based on a linear combination of the features estimated with tract profiling.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{kruper2021evaluating}, which extract
tract profile data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes tract profiles data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

@@

Any images with a b-value less than 100 s/mm\^{}2 were treated as a
\emph{b}=0 image. MP-PCA denoising as implemented in MRtrix3's
\texttt{dwidenoise}\cite{dwidenoise1} was applied with a 5-voxel
window. After MP-PCA, B1 field inhomogeneity was corrected using
\texttt{dwibiascorrect} from MRtrix3 with the N4 algorithm \cite{n4}.
After B1 bias correction, the mean intensity of the DWI series was
adjusted so all the mean intensity of the b=0 images matched across
eachseparate DWI scanning sequence.
\end{itemize}

@@

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was \num{2x2x2}$\text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing pipeline
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox (\texttt{mAFQ})
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Eighteen major
tracts, which are enumerated in the supplemental material, were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalanobis distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

Data from the HBN and Cam-CAN studies were processed using the updated Python
Automated Fiber Quantification toolbox (\texttt{pyAFQ}; \cite{kruper2021evaluating}). In
addition to demonstrating the our analysis pipeline is robust to changes in
tractometry software, the use of the updated \texttt{pyAFQ} capitalized upon
the following improvements over the legacy Matlab version
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item the ability to ingest data provided in the BIDS format
    \cite{gorgolewski2016brain}
    \item the calculation of diffusion kurtosis imaging (DKI
    \cite{jensen2005diffusion}) metrics
\end{enumerate*}
We will refer to the \texttt{mAFQ} and \texttt{pyAFQ} pipeline collectively
as \texttt{AFQ}.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, WA, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, USA
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, WA, USA
\\
\bigskip

@@

samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu},

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance. To do so, we use a linear modeling approach, which aims to
redict phenotypical variance in a group of subjects (or classify group
membership), based on a linear combination of the features estimated with
tract profiling.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.

In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{pyAFQ, kruper2021evaluating}, which extract
tract profile data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes tract profiles data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

@@

Any images with a b-value less than 100 s/mm\^{}2 were treated as a
\emph{b}=0 image. MP-PCA denoising as implemented in MRtrix3's
\texttt{dwidenoise}\cite{dwidenoise1} was applied with a 5-voxel
window. After MP-PCA, B1 field inhomogeneity was corrected using
\texttt{dwibiascorrect} from MRtrix3 with the N4 algorithm \cite{n4}.
After B1 bias correction, the mean intensity of the DWI series was
adjusted so all the mean intensity of the b=0 images matched across
eachseparate DWI scanning sequence.

@@

\item {\it MRtrix3 Reconstruction}

Reconstruction was performed using \emph{QSIprep} 0.12.1. Multi-tissue fiber response functions were estimated using the dhollander algorithm. FODs were estimated via constrained spherical deconvolution (CSD, \mbox{\cite{originalcsd, tournier2008csd}) using an unsupervised multi-tissue method \mbox{\cite{dhollander2019response, dhollander2016unsupervised}}. Reconstruction was done using MRtrix3 \mbox{\cite{mrtrix3}}. FODs were intensity-normalized using mtnormalize \mbox{\cite{mtnormalize}}.}
\end{itemize}

@@

\item Diffusion MRI data from the Cambridge Centre for Ageing and
Neuroscience (Cam-CAN) ``CC700'' dataset
\cite{shafto2014cambridge,taylor2017cambridge}, containing data from 640
subjects with ages 18-88. These data were measured on a 3T Siemens TIM Trio
system and written informed consent was obtained from each participant. Voxel
resolution was \num{2x2x2}$\text{mm}^3$ with 30 non-colinear directions
measured for each of $b=1000$ $\text{s} / \text{mm}^2$ and $b=2000$
$\text{s} / \text{mm}^2$. The diffusion weighted images were acquired
with a twice refocused spin-echo sequence and the same preprocessing and reconstruction pipeline
used for the HBN dataset was applied to this data. We will refer to this
dataset as Cam-CAN.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}, version 1.1 for WH and version 1.2 for ALS)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Eighteen major
tracts, which are enumerated in the supplemental material, were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalanobis distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

Data from the HBN and Cam-CAN studies were processed using the updated Python
Automated Fiber Quantification toolbox (\texttt{pyAFQ}; \cite{pyAFQ}). In
addition to demonstrating the our analysis pipeline is robust to changes in
tractometry software, the use of the updated \texttt{pyAFQ} capitalized upon
the following improvements over the legacy Matlab version
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item the ability to ingest data provided in the BIDS format
    \cite{gorgolewski2016brain}
    \item the calculation of diffusion kurtosis imaging (DKI
    \cite{jensen2005diffusion}) metrics
\end{enumerate*}
We will refer to the \texttt{mAFQ} and \texttt{pyAFQ} pipeline collectively
as \texttt{AFQ}.
